---
title: "Example: Assessing Bayesian Numerical Analysis"
author: "VectorBiTE Training Workshop"
date: "June 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up workspace

```{r, results="hide", warning=FALSE, message=FALSE}
# Set working directory
# setwd('')

# Load libraries
require(R2jags) # does the fitting
require(coda) # makes diagnostic plots
##require(mcmcplots) # another option for diagnostic plots
```

## Numerical Analysis for Bayesian Statistics

Recall the steps to making inference:

- (1) Research question
- (2) Data collection
- (3) Model $Y_i \approx f(X_i)$ 
- (4) Estimate the parameter in the model with uncertainty
- (5) Make inference

To do step 5  we need the following:

### (a) Assess MCMC convergence

MCMC is a convergence algorithm, we reach convergence once we have random draws from the posterior distribution. This does not mean that the draws flat-rise to a single value, but it means that my draws are seemingly unrelated and random. To assess convergence we look at trace plots, the goal is to get "fuzzy caterpillars" looking curves. 


### (b) Summarize draws from MCMC

To summarize draws we create histogram for all draws for each parameter and calculate expectation, variance, credible interval, and so on.

### (c) Prior Sensitivity

We assess prior sensitivity by changing prior values and check whether it affects our results or not. If it does, that means that we are too sensitive to that prior and that is not good for our analysis.

### (d) Make inference

Here, we use the values from part (b) to make inference and answer the research question.

## Example: Midge Wing Length

We will use this simple example to go through the steps of assessing a Bayesian model.

Grogan and Wirth (1981) provide data on the wing length (in millimeters) of nine members of a species of midge (small, two-winged flies). 

From these measurements we wish to make inference about the population mean $\mu$.


## Prior Information

Studies from other populations suggest that wing lengths are usually around 1.9 mm, so we set $\mu_0 = 1.9$

We also know that lengths must be positive ($\mu >0$)

We can approximate this restriction with a normal prior distribution for $\mu$ as follows:

Since most of the normal density is within two standard deviations of the mean we choose $\tau^2_0$ so that

$$ \mu_0 - 2\tau_0 >0 \Rightarrow \tau_0 <1.2/2 = 0.95 $$


To keep it simple we chose for the variance $\sigma^2$ to follow inverse gamma distribution, meaning that $1/\sigma^2$ follows a gamma distribution with $a=b=0.01.$ As a result, our prior is a conjugate prior and the posterior is a normal distribution as well.

## Load data

```{r}
# Load data
WL.data <- read.csv("../data/MidgeWingLength.csv")
Y <- WL.data$WingLength
n <- length(Y)

hist(Y,breaks=20,xlab="Wing Length (mm)") 


mu0 <- 1.9
s20 <- 0.8
a   <- 0.01
b   <- 0.01
```

```{r}
library(rjags)  

model1 <- "model{

  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(mu,inv.var)
  }

  # Prior for mu
  mu        ~ dnorm(mu0,inv.var0)
  inv.var0 <- 1/s20

  # Prior for the inverse variance
  inv.var   ~ dgamma(a, b)

  # Compute the variance
  s2       <- 1/inv.var
}# close model
"

model <- jags.model(textConnection(model1), 
                    n.chains = 3,
                    data = list(Y=Y,n=n,mu0=mu0,s20=s20,a=a,b=b))

update(model, 10000, progress.bar="none"); # Burnin for 10000 samples

samp <- coda.samples(model, 
        variable.names=c("mu","s2"), 
        n.iter=20000, progress.bar="none")


```

## Assessing Convergence by looking at trace plots

```{r}
plot(samp)
```

## Summarize draws from MCMC
```{r}
summary(samp)
```

## Prior Sensitivity
Here you can try different prior parameter and check your results. 

```{r}
mu0 <- 0
s20 <- 10
a   <- 1
b   <- 1
```

```{r}
library(rjags)  
model2 <- "model{

  # Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(mu,inv.var)
  }

  # Prior for mu
  mu        ~ dnorm(mu0,inv.var0)
  inv.var0 <- 1/s20

  # Prior for the inverse variance
  inv.var   ~ dgamma(a, b)

  # Compute the variance
  s2       <- 1/inv.var
}"
```

```{r}
model <- jags.model(textConnection(model2), 
                    n.chains = 3,
                    data = list(Y=Y,n=n,mu0=mu0,s20=s20,a=a,b=b))
```
```{r}
update(model, 10000, progress.bar="none"); # Burnin for 10000 samples

samp <- coda.samples(model, 
        variable.names=c("mu","s2"), 
        n.iter=20000, progress.bar="none")


```

```{r}
plot(samp)
```

```{r}
summary(samp)
```

## Make Inference

Here we use values from our summary to answer our research question. eg: Given the data, we predict that midge wing length is $E(\mu|\tilde{Y}) =1.8mm$ with variance $V(\mu|\tilde{Y}) =0.05^2.$