---
title: 'VectorBiTE Training Workshop: <br>Dengue Challenge Solution'
author: ''
date: "June 2018"
output:
  html_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Dengue Incidence Data


The data for this question consist of time series of dengue case data from San Juan, Puerto Rico together with environmental data for each location across a number of transmission seasons. The data files are in the file  ${ {\tt combined\_sanjuan\_new.csv}$. Detailed descriptions of the data are available at ${\tt http://dengueforecasting.noaa.gov/}$. 

To operate effectively, health departments must be able to predict weekly cases, as this will correspond to hospital demand and resources.  Your task is to provide a fitted model for forecasting weekly total dengue cases in San Juan. You can use autoregressive components, sine/cosine, trends to build your model. You can also use the environmental covariates. Remember that you want to be able to predict into the future so you will only include lagged predictors into your prediction model. Below we suggest a series of steps for the analysis and then you'll have the opportunity to develop your own model(s) and choose which components that you want to keep. We'll compare models during the group discussion.  

First you should read in the data

```{r sjdat}
### load in the data
sanjuan<-read.csv(file="../data/combined_sanjuan_new.csv")
names(sanjuan)
summary(sanjuan)
```

Next let's plot the target output, ${\tt total_cases}$ across time:

```{r sj1, fig.width=6, fig.height=4}
t<-seq(1, length(sanjuan$total_cases))

plot(t, sanjuan$total_cases, type="l")
```

It may be a little hard to tell, but the variance when you have a lot of cases the variance is higher, too. As in the airline example we want to do a transformation. However, we have zeros in the total cases, so instead of a log we'll use a square-root.

```{r, fig.width=6, fig.height=4}
t<-seq(1, length(sanjuan$total_cases))

plot(t, sqrt(sanjuan$total_cases), type="l")
```

Next, look at the ACF plot for the square-root of total cases
```{r}
acf(sanjuan$total_cases, lag.max=52)
```

So we'll definitely want to include autocorrelation. 

If you were to look at your covariates, you'd notice that there's a lot of correlation between things like average and max or min temperature and population and the adjusted population. So we don't want to include absolutely everything here, so we're going to select things down a bit. 

First, you want to build a new data frame with a subset of possible covariates. We start with the sqrt of cases as your response, AR1 of the sqrt response, a trend, and sine/cosine with a 52 week period. 
```{r}
n<-max(t)
YX <- data.frame(sqrty=sqrt(sanjuan$total_cases)[2:n],
                 sqrty.m1=sqrt(sanjuan$total_cases)[1:(n-1)],
                 t=t[2:n],
                 sin1=sin((2:n)*2*pi/52),
                 cos1=cos((2:n)*2*pi/52),
                 season=sanjuan$season[2:n],
                 w=sanjuan$season_week[2:n],
                 lpop.m1=log(sanjuan$adjpop+1)[1:(n-1)],
                 lp.m1=log(sanjuan$prec+1)[1:(n-1)],
                 tavg.m1=sanjuan$tavg[1:(n-1)],
                 ndvi45.m1=sanjuan$NDVI.18.45.66.14.[1:(n-1)],
                 ndvi50.m1=sanjuan$NDVI.18.50..66.14.[1:(n-1)],
                 nino12.m1=sanjuan$nino12[1:(n-1)],
                 soi.m1=sanjuan$soi[1:(n-1)])
```

```{r}
pairs(YX[c(4,5, 8:14)])

```


Now fit a linear model with the trend, and sine/cosine components, only.
```{r}
mod1 <- lm(sqrty ~ t + sin1 + cos1, data=YX)
summary(mod1)
```

___Q3: Are all of the predictors significant? What is the $R^2$ of your regression?___


Next, we plot the data on a square-root scale (drawn with a dotted line).  Overlay the FITTED values from your regression (e.g., if your regression model was called "reg" you want to plot reg$fitted vs t) as a solid line in another color. This solid line is your prediction.

```{r, fig.align="center", fig.width=6, fig.height=4}
plot(YX$t, YX$sqrty, xlab="season",
     ylab="sqrt weekly cases", type="l", col=4, lty=2,
     xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
lines(YX$t, mod1$fitted, col=2, lwd=2)
legend("topleft", legend=c("data", "fitted"), lty=c(2,1), col=c(4,2))
```
The difference between the solid and dotted lines at each month are your residuals across time. As always, we want to also look at our residuals explicitly to see if we're doing a good job of explaining things. For TS we primarily look at residuals across time, and the ACF of our residuals. So make those two plots here. 

```{r, fig.align="center", fig.width=6, fig.height=4}
par(mfrow=c(1,2))
plot(mod1$resid, xlab="year", ylab="residual", type="l",
     col=4, main="residuals in time",  xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
acf(mod1$resid, lwd=2, lag.max=55)
```
We're missing a lot. 

Now add in the AR component and check things again.

```{r}
mod2 <- lm(sqrty ~ sqrty.m1 + t + sin1 + cos1, data=YX)
summary(mod2)
```


```{r, fig.align="center", fig.width=6, fig.height=4}
plot(YX$t, YX$sqrty, xlab="season",
     ylab="sqrt weekly cases", type="l", col=4, lty=2,
     xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
lines(YX$t, mod2$fitted, col=2, lwd=2)
legend("topleft", legend=c("data", "fitted"), lty=c(2,1), col=c(4,2))
```


```{r, fig.align="center", fig.width=6, fig.height=4}
par(mfrow=c(1,2))
plot(mod2$resid, xlab="year", ylab="residual", type="l",
     col=4, main="residuals in time",  xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
acf(mod2$resid, lwd=2, lag.max=55)
```
There's not a lot that's obvious that's going on here. Let's try the boxplots of residuals by week in the season and see if there's anything obvious.

```{r, fig.align="center", fig.width=6, fig.height=4}
boxplot(mod2$resid ~ YX$w[YX$t], xlab="week",
        ylab="residuals", col=7)
```

Not a lot obviously going on. By trial and error (or if you know another way, such as the step function) try to build a better model while staying parsimonious. 

```{r}
mod3 <- lm(sqrty ~ sqrty.m1 + t + sin1 + cos1 + nino12.m1, data=YX)
summary(mod3)
```


```{r, fig.align="center", fig.width=6, fig.height=4}
plot(YX$t, YX$sqrty, xlab="season",
     ylab="sqrt weekly cases", type="l", col=4, lty=2,
     xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
lines(YX$t, mod3$fitted, col=2, lwd=2)
legend("topleft", legend=c("data", "fitted"), lty=c(2,1), col=c(4,2))
```


```{r, fig.align="center", fig.width=6, fig.height=4}
par(mfrow=c(1,2))
plot(mod3$resid, xlab="year", ylab="residual", type="l",
     col=4, main="residuals in time",  xaxt="n", lwd=2)
axis(1, at=(0:18)*52, labels=levels(YX$season))
acf(mod3$resid, lwd=2, lag.max=55)
```

Now compare the 3 models via BIC. Calculate the relative model probabilities.
```{r}
n<-length(YX$sqrty)-1
bics<-c(mod1=extractAIC(mod1, k=log(n))[2],
        mod2=extractAIC(mod2, k=log(n))[2],
        mod3=extractAIC(mod3, k=log(n))[2])

ebics<-exp(-0.5*(bics-min(bics)))

probs<-ebics/sum(ebics)

round(probs, 5)
```


## Extra challenge

For convenience I simply chose a lag of 1 week back when I built the data set. However, we know that the lad time between an infected human and another infected human (because of the various incubation and development times in the human and the mosquito) is likely to be weeks. So we expect there to be longer lags. We can use the ccf (for cross-correlation) function to see how covariates relate to the response at different lags. For instance:
```{r}
ccf(sanjuan$tavg, sanjuan$total_cases)
```

This indicates that the temperature about 10 weeks back is more correlated with the incidence than the temperature 1 week ago. 

Plot a few other ccfs, choose some alternative lags, and build a new dataset with these lagged variables.

```{r, fig.height=10}
ws<-c(16, 22, 23, 24)
ws.l<-c(20)
par(mfrow=c(3,2))
for(i in 1:4){
  ccf(sanjuan[,ws[i]], sanjuan$total_cases)
}
ccf(log(sanjuan[,ws.l]+1), sanjuan$total_cases)
names(sanjuan[,c(ws, ws.l)])
```

So it looks like 
NDVI.18.50..66.14.: -10
tavg: -10
nino12: -5
nino34: -12
logprec: -5  

Since the el nino indices are correlated I'll just keep the first, and I'll build a new dataset:
```{r}
n<-max(t)
l<-10
ts<-(l+1):n
YX2 <- data.frame(sqrty=sqrt(sanjuan$total_cases)[ts],
                 sqrty.m1=sqrt(sanjuan$total_cases)[ts-1],
                 t=ts,
                 sin1=sin((ts)*2*pi/52),
                 cos1=cos((ts)*2*pi/52),
                 season=sanjuan$season[ts],
                 w=sanjuan$season_week[ts],
                 lp.m5=log(sanjuan$prec+1)[ts-5],
                 tavg.m10=sanjuan$tavg[ts-10],
                 tavg2.m10=(sanjuan$tavg[ts-10])^2,
                 ndvi50.m10=sanjuan$NDVI.18.50..66.14.[ts-10],
                 nino12.m5=sanjuan$nino12[ts-5])
```

```{r}
mod4 <- lm(sqrty ~ tavg2.m10 + t+sin1+sqrty.m1, data=YX2)
summary(mod4)
```

```{r}
mod5 <- lm(sqrty ~ . - season - sin1 - cos1 - t - nino12.m5 -lp.m5-ndvi50.m10, data=YX2)
summary(mod5)
```

```{r}
mod6 <- lm(sqrty ~ sin1 + cos1 + t +sqrty.m1, data=YX2)
summary(mod6)
```

```{r}
n<-length(YX2$sqrty)-1
bics<-c(mod4=extractAIC(mod4, k=log(n))[2],
        mod5=extractAIC(mod5, k=log(n))[2],
        mod6=extractAIC(mod6, k=log(n))[2])

ebics<-exp(-0.5*(bics-min(bics)))

probs<-ebics/sum(ebics)

round(probs, 5)
```
<br>
<br>
<br>