---
title: 'CMDA/STAT 4664: <br> Homework 1: Solution'
author: "Leah R Johnson"
date: "February 2017"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  html_document: default
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Albatross chick growth data

```{r alby1, fig.width=4, fig.height=4}
### load in the data
alb<-read.csv(file="../data/albatross_grow.csv")
alb<-subset(x=alb, !is.na(alb$wt))
plot(alb$age, alb$wt, xlab="age (days)", ylab="weight (g)", xlim=c(0, 100))

```

### Von Bert Model

In class we defined the Von Bert model primarily in terms of lengths, but here we want weight. So recall that we defined $W=\rho L^3$. Using the results from class and your exercises we can write:
$$
W(t) = \rho (L_{\infty}(1-e^{-Kt})+L_0 e^{-Kt})^3
$$
If I pull out $L_{\infty}$ and define $c=L_0/L_{\infty}$ and $W_{\infty}=\rho L_{\infty}^3$ this equation becomes:
$$
W(t) = W_{\infty}(1-e^{-Kt}+ c e^{-Kt})^3.
$$
$W_{\infty}$ is interpretted as the mean asymptotic weight, and $c$ the ratio between the initial and final lengths. This second equation is the one I will fit.

### Fitting three curves using NLS

First I will define my 2 bespoke functions (I'll use the lm function for the line, as it fits using least squares automatically). 

```{r}
logistic1<-function(t, r, K, N0){
  N0*K*exp(r*t)/(K+N0*(exp(r*t)-1))
}

vonbert.w<-function(t, Winf, c, K){
  Winf*(1 - exp(-K*t) + c*exp(-K*t))^3
}

```

Now I fit all 3 models using least squares. As for the logistic example in class, I'll fit the scaled data, to improve the stability of my estimates.

```{r}
scale<-4000

alb.lin<-lm(wt/scale~age, data=alb)

alb.log<-nls(wt/scale~logistic1(age, r, K, N0), 
          start=list(K=1, r=0.1, N0=0.1), data=alb)

alb.vb<-nls(wt/scale~vonbert.w(age, Winf, c, K), 
          start=list(Winf=0.75, c=0.01, K=0.01), data=alb)

```

Next I will build predictions for each of my models at a new set of ages

```{r}
ages<-seq(0, 100, length=1000)

pred.lin<-predict(alb.lin, newdata = list(age=ages))*scale

pred.log<-predict(alb.log, newdata = list(age=ages))*scale

pred.vb<-predict(alb.vb, newdata = list(age=ages))*scale

```

And finally plot the data with the fits
```{r alb_fits, fig.width=6, fig.height=4}

plot(alb$age, alb$wt, xlab="age (days)", ylab="weight (g)", xlim=c(0,100))
lines(ages, pred.lin, col=2, lwd=2)
lines(ages, pred.log, col=3, lwd=2)
lines(ages, pred.vb, col=4, lwd=2)

legend("topleft", legend = c("linear", "logistic", "Von Bert"), lwd=2, lty=1, col=2:4)

```

Next I examine the residuals between the 3 models:
```{r alb_resid, fig.width=4, fig.height=8}

par(mfrow=c(3,1), bty="n")
plot(alb$age, resid(alb.lin), main="LM resids", xlim=c(0,100))
plot(alb$age, resid(alb.log), main="Logisitic resids", xlim=c(0,100))
plot(alb$age, resid(alb.vb), main="VB resids", xlim=c(0,100))


```

The residuals for all 3 models still exhibit some patterns. In particular, the data seems to go down near the end of the observation period, but none of these models can capture that behavior. 

Finally, let's compare the 3 models:
```{r}
n<-length(alb$wt)
list(lin=signif(sum(resid(alb.lin)^2)/(n-2*2), 3), 
    log= signif(sum(resid(alb.log)^2)/(n-2*3), 3), 
    vb= signif(sum(resid(alb.vb)^2)/(n-2*3), 3))
```

The logistic model has the lowest adjusted SSE, so it's the best by this measure. It is also, visually, a better fit. 

## Aedes aegypti fecundity data

```{r aedes1, fig.width=5, fig.height=4}
### load in the data
aedes<-read.csv(file="../data/aedes_fecund.csv")

plot(aedes$T, aedes$EFD, xlab="temperature (C)", ylab="Eggs/day")

```


### Fitting three curves using NLS

I define my own quadratic function (to make it easier to choose starting values) and so that I can force the function to be equal to zero above and below the minimum and maximum temperature thresholds, as well as the Briere function.

```{r}
quad1<-function(T, T0, Tm, c){
  c*(T-T0)*(T-Tm)*as.numeric(T<Tm)*as.numeric(T>T0)
}

briere<-function(T, T0, Tm, c){
  c*T*(T-T0)*(abs(Tm-T)^(1/2))*as.numeric(T<Tm)*as.numeric(T>T0)
}
```

Now I fit all 3 models using least squares. Although it's not as necessary here (as the data aren't as large as the albatross example), I will scale the data. 

```{r}
scale<-20

aed.lin<-lm(EFD/scale~T, data=aedes)

aed.quad<-nls(EFD/scale~quad1(T, T0, Tm, c), 
          start=list(T0=10, Tm=40, c=0.01), data=aedes)

aed.br<-nls(EFD/scale~briere(T, T0, Tm, c), 
          start=list(T0=10, Tm=40, c=0.1), data=aedes)

```

Then I build predictions for each of my models at a new set of temperatures.

```{r}
temps<-seq(0, 40, length=1000)

pred.lin<-predict(aed.lin, newdata = list(T=temps))*scale

pred.quad<-predict(aed.quad, newdata = list(T=temps))*scale

pred.br<-predict(aed.br, newdata = list(T=temps))*scale

```

And finally plot the data with the fits
```{r aedes_fits, fig.width=6, fig.height=5}

plot(aedes$T, aedes$EFD, xlab="Temp (C)", ylab="Eggs/Day", xlim=c(5,40))
lines(temps, pred.lin, col=2, lwd=2)
lines(temps, pred.quad, col=3, lwd=2)
lines(temps, pred.br, col=4, lwd=2)

legend("topleft", legend = c("linear", "quadratic", "Briere"), lwd=2, lty=1, col=2:4)

```

Next I examine the residuals between the 3 models:
```{r aedes_resid, fig.width=6, fig.height=3.5}

par(mfrow=c(1,3), bty="n")
plot(aedes$T, resid(aed.lin), main="LM resids", xlim=c(10,40))
plot(aedes$T, resid(aed.quad), main="quad resids", xlim=c(10,40))
plot(aedes$T, resid(aed.br), main="Briere resids", xlim=c(10,40))


```

Finally, let's compare the 3 models:
```{r}
n<-length(aedes$T)
list(lin=signif(sum(resid(aed.lin)^2)/(n-2*2), 3), 
    quad= signif(sum(resid(aed.quad)^2)/(n-2*3), 3), 
    br= signif(sum(resid(aed.br)^2)/(n-2*3), 3))
```

Surprisingly (to me) the quadratic function comes out on top here. Note that if I hadn't fit the quadratic function with the extra constraints that be zero if $T<T_0$ of $T>T_m$, the Briere would come out on top (and, really, all are very similarly bad!):

```{r}
quad2<-function(T, T0, Tm, c){
  c*(T-T0)*(T-Tm)
}

scale<-20
aed.quad2<-nls(EFD/scale~quad2(T, T0, Tm, c), 
          start=list(T0=10, Tm=40, c=0.01), data=aedes)

pred.quad2<-predict(aed.quad2, newdata = list(T=temps))*scale

n<-length(aedes$T)
list(lin=signif(sum(resid(aed.lin)^2)/(n-2*2), 4), 
    quad= signif(sum(resid(aed.quad)^2)/(n-2*3), 4), 
    br= signif(sum(resid(aed.br)^2)/(n-2*3), 4), 
    quad2= signif(sum(resid(aed.quad2)^2)/(n-2*3), 4))

```

```{r aedes_fits2, fig.width=6, fig.height=5}

plot(aedes$T, aedes$EFD, xlab="Temp (C)", ylab="Eggs/Day", xlim=c(5,40))
lines(temps, pred.lin, col=2, lwd=2)
lines(temps, pred.quad, col=3, lwd=2)
lines(temps, pred.br, col=4, lwd=2)
lines(temps, pred.quad2, col=5, lwd=2)

legend("topleft", legend = c("linear", "quadratic", "Briere", "quadratic2"), lwd=2, lty=1, col=2:5)

```



<br>
<br>
<br>