---
title: 'Example: Fitting Traits with JAGS (Bayesian Inference)'
author: 'VectorBiTE Training Workshop'
date: 'June 2018'
output:
  html_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Set up workspace

```{r, results="hide", warning=FALSE, message=FALSE}
# Set working directory
# setwd('')

# Load libraries
require(R2jags) # does the fitting
require(coda) # makes diagnostic plots
##require(mcmcplots) # another option for diagnostic plots
```

```{r}
# Load data
dat <- read.csv("../data/AeaegyptiTraitData.csv")
```

## Data exploration and preparation

These data are traits from _Ae. aegypti_ mosquitoes measured across temperature in lab experiments. The traits we have data for are:  
- pEA: proportion surviving from egg to adulthood  
- MDR: mosquito development rate  
- PDR: parasite development rate (= 1/EIP the extrinsic incubation period)  
- $\mu$ (mu): death rate (= 1/longevity)  
Note that some of the traits come in multiple forms (e.g., $\mu$ and 1/$\mu$, PDR and EIP).

### Inspect the data set
```{r}
# Look at the data (just first 3 columns, first 15 rows)
dat[1:15,1:3]
```

### Subset and plot the $\mu$ and lifespan data
For this example we'll fit a thermal response for adult mosquito mortality. Some relevant data was reported as lifespan and some  was reported $\mu$ (= 1/lifespan). In the plots below, note that the $\mu$ data is u-shaped and the lifespan data is unimodal (hump-shaped).  

```{r, fig.align='center', fig.height=4, fig.width=7}
dat.mu <- subset(dat, trait.name == "mu")
dat.lf <- subset(dat, trait.name == "1/mu")

par(mfrow=c(1,2), bty="l") 
plot(trait ~ T, data = dat.mu, ylab="mu")
plot(trait ~ T, data = dat.lf, ylab="1/mu")
```

### Combine $\mu$ and lifespan data and replot them together
Since thermal biology theory is based on unimodal thermal responses, we want to fit the thermal response to the lifespan data format. Thus, we'll need to convert the $\mu$ data to lifespan by taking the inverse. The combined data should have a nice unimodal shape that we can fit a function to.

```{r, fig.align='center', fig.height=4, fig.width=5}
dat.mu.inv <- dat.mu # make a copy of the mu data
dat.mu.inv$trait <- 1/dat.mu$trait # take the inverse of the trait values to convert mu to lifespan
dat.lf.comb <- rbind(dat.mu.inv, dat.lf) # combine both lifespan data sets together 
  
plot(trait ~ T, data = dat.lf.comb, ylab="Lifespan")
```

## Fitting a trait thermal response curve using JAGS

### The two thermal responses
Most thermal response curves can be reasonably fit using one of two thermal reponses: traits that respond symmetrically to temperature are fit with a quadratic function and traits that respond assymetrically to temperature are fit with a Briere function. In both functions, T0 is the lower thermal limit and Tm is the upper thermal limit (i.e., where the trait value goes to zero on either end) and q scales the value at the optimum.   

Quadratic function (re-arranged from the typical formulation so that parameters are x-intercepts T0 and Tm): 
$$
f(T) =-q (T-T0) (T-Tm)
$$
Briere function: 
$$
f(T) =q T (T-T0)\sqrt{(Tm-T)}
$$

### The thermal response model file
JAGS needs the model written as a .txt/.bug file inside your the working directory. You can either make the text file directly, or create it using the sink() function via your R script (see below). The model file has two mandatory sections (the priors and the likelihood) and one optional section (derived quantiaties calculated from your fitted parameters).  

In the example below for a quadratic function, the priors are specified by uniform distributions (the two arguments specify the lower and upper bounds, respectively). Note that unlike R and most other programs, JAGS uses $\tau$ instead of $\sigma$ to parameterize the variance of the normal distribution ($\tau = 1/(\sigma^2)$). However it can be easier to specify the prior on sigma directly.

The likelihood below can be interpreted as follows: the observed data are normally distributed where the mean across temperature follows the parameterized quadratic function. We use the inequalities so that the trait mean goes to zero when T < T0 and T > Tm, instead of going negative.  

The derived quantity section below is used to calculate the trait value over a temperature gradient at the same time as the parameters are being fitted. You'll notice that it's simply the quadratic function from the likelihood function restated where temperature varies over a gradient instead of according to the data points. The derived trait name starts with 'z' so alphabetically it will beat the bottom of the output, and all our parameters will be at the top. 

Calculating the dervied quanitity here (as opposed to afterward from the parameters) has two benefits: 1) it makes it easier to plot the thermal response and 2) it makes it easier to pull out the traits values for use in downstream analyses (e.g., calculating $R_0$.)
```{r eval = FALSE}
  
sink("quad.txt")
cat("
    model{
    
    ## Priors
    cf.q ~ dunif(0, 1)
    cf.T0 ~ dunif(0, 24)
    cf.Tm ~ dunif(25, 45)
    cf.sigma ~ dunif(0, 1000)
    cf.tau <- 1 / (cf.sigma * cf.sigma)
    
    ## Likelihood
    for(i in 1:N.obs){
    trait.mu[i] <- -1 * cf.q * (temp[i] - cf.T0) * (temp[i] - cf.Tm) * (cf.Tm > temp[i]) * (cf.T0 < temp[i])
    trait[i] ~ dnorm(trait.mu[i], cf.tau)
    }
    
    ## Derived Quantities
    for(i in 1:N.Temp.xs){
    z.trait.mu.pred[i] <- -1 * cf.q * (Temp.xs[i] - cf.T0) * (Temp.xs[i] - cf.Tm) * (cf.Tm > Temp.xs[i]) * (cf.T0 < Temp.xs[i])
    }
    
    } # close model
    ",fill=T)
sink()
```

### Additional settings for JAGS 
```{r}
# Parameters to Estimate
parameters <- c("cf.q", "cf.T0", "cf.Tm","cf.sigma", "z.trait.mu.pred")

# Initial values for the parameters
inits<-function(){list(
  cf.q = 0.01,
  cf.Tm = 35,
  cf.T0 = 5,
  cf.sigma = rlnorm(1))}

# MCMC Settings: number of posterior dist elements = [(ni - nb) / nt ] * nc
ni <- 25000 # number of iterations in each chain
nb <- 5000 # number of 'burn in' iterations to discard
nt <- 8 # thinning rate - jags saves every nt iterations in each chain
nc <- 3 # number of chains

# Temperature sequence for derived quantity calculations
Temp.xs <- seq(0, 45, 0.2)
N.Temp.xs <-length(Temp.xs)
```

### Fitting the trait thermal response

```{r, results="hide"}
# Pull out data columns as vectors
data <- dat.lf.comb # this lets us reuse the same generic code: we only change this first line
trait <- data$trait
N.obs <- length(trait)
temp <- data$T

# Bundle all data in a list for JAGS
jag.data<-list(trait = trait, N.obs = N.obs, temp = temp, Temp.xs = Temp.xs, N.Temp.xs = N.Temp.xs)

# Run JAGS
lf.fit <- jags(data=jag.data, inits=inits, parameters.to.save=parameters, 
               model.file="quad.txt", n.thin=nt, n.chains=nc, n.burnin=nb, 
               n.iter=ni, DIC=T, working.directory=getwd())

## change into "mcmc" type samples for visualization with CODA
lf.fit.mcmc<-as.mcmc(lf.fit)

```

### Running diagnostics

```{r}
## View the parameters - only the first 5 lines (4 parameters + the devience) 
## otherwise it will also show you all of your derived quantities (i.e., the trait value over the entire temperature gradient at 0.2C increments)
lf.fit$BUGSoutput$summary[1:5,]
```

```{r, fig.align='center', fig.height=8}
## plot the chains using the coda package
plot(lf.fit.mcmc[,c(1,3,4)])
```

### Plot thermal response

We can easily plot the mean thermal response and the CIs using the derived quantities that we calculated from the fitted parameters.  

```{r, fig.align='center', fig.height=4.5, fig.width=6}
plot(trait ~ T, xlim = c(0, 45), ylim = c(0,42), data = dat.lf.comb, ylab = "Lifespan for Ae. aegypti", xlab = "Temperature")
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "2.5%"] ~ Temp.xs, lty = 2)
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "97.5%"] ~ Temp.xs, lty = 2)
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "mean"] ~ Temp.xs)
```

### Additional output

We can use the which.max() function to find the optimal temperature for adult lifespan.

We can pull out the lifespan values for each iteration of the MCMC chain over the temperature gradient to calculate $R_0$ later on. 

```{r}
# Calculate the optimal temperature for lifespan
Temp.xs[which.max(as.vector(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "mean"]))]

# Pull out the trait values over the temperature gradient for later use
lf.grad <- lf.fit$BUGSoutput$sims.list$z.trait.mu.pred
dim(lf.grad) # A matrix with 7500 iterations of the MCMC chains at 226 temperatures
```