---
title: 'VectorBiTE Training Workshop: <br>Traits + MLE Challenge Problem'
author: ''
date: "June 2018"
output:
  html_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Mosquito Thermal Trait Data

The data for this question consist thermal traits for _Aedes agypti_ mosquitos found in the data file ${\tt AeaegyptiTraitData.csv}$. There are four possible traits (as well as some data that are recorded as the inverse trait, which we ignore for now):

- pEA: proportion surviving from egg to adulthood  
- MDR: mosquito development rate  
- PDR: parasite development rate (= 1/EIP the extrinsic incubation period)  
- $\mu$ (mu): death rate (= 1/longevity)

You will choose one of these four traits and fit a curve to the data by maximum likelihood. (You can also fit a curve to the data of your choosing.) You will base your code on that provided in the lecture/practical portion of the class. We provide a specific set of tasks to accomplish starting with easy tasks through to more challenging aspects. We'll compare results and models during the group discussion.

First we'll read in the data and look at the summary

```{r sjdat}
### load in the data
dat<-read.csv(file="../data/AeaegyptiTraitData.csv")
summary(dat)
```

Next let's plot the response output, ${\tt total\_cases}$ across time:

```{r sj1, fig.width=6, fig.height=6}
traits<-c("pEA", "MDR", "PDR", "mu")
par(mfrow=c(2,2), bty="o", mai=c(0.8,0.7,0.1,0.1))
for(i in 1:4){
  d.temp<-subset(dat, trait.name==traits[i])
  plot(trait ~ T, data=d.temp, xlab="Temperature (C)", ylab=traits[i], xlim=c(5, 45))
}

```



### Task 1: Fitting a line, revisited

1. Using the _nll.slr_ function and the code from the lecture, find the MLEs of the slope and intercept (i.e., find the best fitting line for your chosen trait). 
2. Plot your data with the fitted line.
3. Plot the likelihood surface for $b_0$ and $b_1$ and indicate the MLEs your likelihood surface. 
4. Obtain confidence intervals for your estimates. 


### Task 2: Fitting a non-linear trait model

1. Choose a non-linear model to fit to your trait (e.g., quadratic, Briere...). We provide implemented functions in the code file ${\tt temp\_functions.R}$. 
2. Using the _nll.slr_ function as an example, write your own function that calculates the negative log likelihood as a function of the parameters describing your trait and any additional parameters you need for an appropriate noise distribution (e.g., $\sigma$ if you have normal noise).
3. Use the _optim_ function to find the MLEs of all of your parameters.
4. Obtain a confidence interval for your estimate.
5. Plot the fitted function with the data and your fittend line from Task 1.


### Task 2b: Fitting a second non-linear trait (optional)

Choose a second non-linear function to fit to your trait data. Find the MLEs and fit this curve to your data. Plot the fit together with your previous fits and the data



### Task 3: Compare models

Now compare your 2-3 models via AIC or BIC (optional: calculate the relative model probabilities). Which comes out on top? Is this what you expected?


### Extra challenge

If you have the time and the interest you may try to fit curves to all of the 4 traits!

<br>
<br>
<br>