---
title: 'VectorBiTE Training Workshop: Fitting Traits with JAGS (Bayesian Inference)'
author: ''
date: 'June 2018'
output:
  html_document: default
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Set up workspace

```{r, results="hide", warning=FALSE, message=FALSE}
# Set working directory
# setwd('')

# Load libraries
require(R2jags) # does the fitting
require(coda) # makes diagnostic plots
##require(mcmcplots) # another option for diagnostic plots
```

```{r}
# Load data
Aaeg.data <- read.csv("../data/AeaegyptiTraitData.csv")
```

## Data exploration and preparation

These data are traits from _Ae. aegypti_ mosquitoes measured across temperature in lab experiments. The traits we have data for are:  
- pEA: proportion surviving from egg to adulthood  
- MDR: mosquito development rate  
- PDR: parasite development rate (= 1/EIP the extrinsic incubation period)  
- $\mu$ (mu): death rate (= 1/longevity)  
Note that some of the traits come in multiple forms (e.g., $\mu$ and 1/$\mu$, PDR and EIP).

### Inspect the data set
```{r}
# Look at the data (just first 3 columns, first 15 rows)
Aaeg.data[1:15,1:3]
```

### Subset and plot the $\mu$ and lifespan data

Note that the $\mu$ data is u-shaped and the lifespan data is unimodal (hump-shaped).  

```{r, fig.align='center', fig.height=4, fig.width=7}
mu.data <- subset(Aaeg.data, trait.name == "mu")
lf.data <- subset(Aaeg.data, trait.name == "1/mu")

par(mfrow=c(1,2), bty="l") 
plot(trait ~ T, data = mu.data, ylab="mu")
plot(trait ~ T, data = lf.data, ylab="1/mu")
```

### Combine $\mu$ and lifespan data and replot them together
Since thermal biology theory is based on unimodal thermal responses, we want to fit the trait as lifespan instead of $\mu$. Thus, we'll need to convert the $\mu$ data to lifespan by taking the inverse. The combined data should have a nice unimodal shape that we can fit a function to.

```{r, fig.align='center', fig.height=4, fig.width=5}
mu.data.inv <- mu.data # make a copy of the mu data
mu.data.inv$trait <- 1/mu.data$trait # take the inverse of the trait values to convert mu to lifespan
lf.data.comb <- rbind(mu.data.inv, lf.data) # combine both lifespan data sets together 
  
plot(trait ~ T, data = lf.data.comb, ylab="1/mu")
```

## Fitting a trait thermal response curve using JAGS

### The two thermal responses
Most thermal response curves can be reasonably fit using one of two thermal reponses: traits that respond symmetrically to temperature are fit with a quadratic function and traits that respond assymetrically to temperature are fit with a Briere function. In both functions, T0 is the lower thermal limit and Tm is the upper thermal limit (i.e., where the trait value goes to zero on either end) and q scales the value at the optimum.   

Quadratic function: -1 q (T-T0) (T-Tm)  
Briere function: q T (T-T0) sqrt(Tm-T)

### The thermal response model file
JAGS needs the model written as a .txt file inside your the working directory. You can either make the text file directly, or create it using the sink() function via your R script (see below).  
The model file has two mandatory sections (the priors and the likelihood) and one optional section (derived quantiaties calculated from your fitted parameters).  

In the example below for a quadratic function, most of the priors are specified via uniform distributions (the two arguments specific the lower and upper bounds, respectively). Note that unlike in R and most other programs, in JAGS the normal distribution uses $\tau$ instead of $\sigma$ in the parameterization for the variance of the normal distribution ($\tau = 1/(\sigma^2)$).  

The likelihood for can be interpreted as follows: the observed data are normally distributed where the mean at a given temperature follows the quadratic equation 
```{r eval = FALSE}
  
sink("quad.txt")
cat("
    model{
    
    ## Priors
    cf.q ~ dunif(0, 1)
    cf.T0 ~ dunif(0, 24)
    cf.Tm ~ dunif(25, 45)
    cf.sigma ~ dunif(0, 1000)
    cf.tau <- 1 / (cf.sigma * cf.sigma)
    
    ## Likelihood
    for(i in 1:N.obs){
    trait.mu[i] <- -1 * cf.q * (temp[i] - cf.T0) * (temp[i] - cf.Tm) * (cf.Tm > temp[i]) * (cf.T0 < temp[i])
    trait[i] ~ dnorm(trait.mu[i], cf.tau)
    }
    
    ## Derived Quantities and Predictions
    for(i in 1:N.Temp.xs){
    z.trait.mu.pred[i] <- -1 * cf.q * (Temp.xs[i] - cf.T0) * (Temp.xs[i] - cf.Tm) * (cf.Tm > Temp.xs[i]) * (cf.T0 < Temp.xs[i])
    }
    
    } # close model
    ",fill=T)
sink()
```

### Additional settings for JAGS 
```{r}
# Parameters to Estimate
parameters <- c("cf.q", "cf.T0", "cf.Tm","cf.sigma", "z.trait.mu.pred")

# Initial values for the parameters
inits<-function(){list(
  cf.q = 0.01,
  cf.Tm = 35,
  cf.T0 = 5,
  cf.sigma = rlnorm(1))}

# MCMC Settings: number of posterior dist elements = [(ni - nb) / nt ] * nc
ni <- 25000 # number of iterations in each chain
nb <- 5000 # number of 'burn in' iterations to discard
nt <- 8 # thinning rate - jags saves every nt iterations in each chain
nc <- 3 # number of chains

# Temperature sequence for derived quantity calculations
Temp.xs <- seq(0, 45, 0.2)
N.Temp.xs <-length(Temp.xs)
```

### Fitting the trait thermal response
```{r, results="hide"}
# Pull out data columns as vectors
data <- lf.data.comb # this lets us reuse the same generic code: we only change this first line
trait <- data$trait
N.obs <- length(trait)
temp <- data$T

# Bundle all data in a list for JAGS
jag.data<-list(trait = trait, N.obs = N.obs, temp = temp, Temp.xs = Temp.xs, N.Temp.xs = N.Temp.xs)

# Run JAGS
lf.fit <- jags(data=jag.data, inits=inits, parameters.to.save=parameters, 
               model.file="quad.txt", n.thin=nt, n.chains=nc, n.burnin=nb, 
               n.iter=ni, DIC=T, working.directory=getwd())

## change into "mcmc" type samples for visualization with CODA
lf.fit.mcmc<-as.mcmc(lf.fit)

```

### Running diagnostics

```{r}
## View the parameters
## only the first 5 lines (otherwise it will also show you all of your derived quantities)
lf.fit$BUGSoutput$summary[1:5,]
```

```{r, fig.align='center', fig.height=8}
## plot the chains using the coda package
plot(lf.fit.mcmc[,c(1,3,4)])
```

```{r}
# mcmcplot will open an html file in your default browse and allow you to look at the chains for convergence and autocorrelation
# mcmcplot(lf.fit)
```

### Plot thermal response

```{r, fig.align='center', fig.height=4.5, fig.width=6}
plot(trait ~ T, xlim = c(0, 45), ylim = c(0,42), data = lf.data.comb, ylab = "Lifespan for Ae. aegypti", xlab = "Temperature")
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "2.5%"] ~ Temp.xs, lty = 2)
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "97.5%"] ~ Temp.xs, lty = 2)
lines(lf.fit$BUGSoutput$summary[6:(6 + N.Temp.xs - 1), "mean"] ~ Temp.xs)
```